Biodiversity temporal trends are reshaping food web structure and redundancy in riverine ecosystems

### 2025-04-16 README.Rmd

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-",
  eval = TRUE 
)

mypath <- rprojroot::find_package_root_file
source(mypath("R", "misc.R"))
data_dir <- mypath("data")

library(tidyverse)
library(magrittr)
library(jmotif)
```
# Analyse desassembly and assembly processes in fish communities

<!-- badges: start -->
<!-- badges: end -->

The goal of this project is to analyze the links between the biomass dynamics
and community structure dynamics.

## Data

data about communities are located in `data` folder.

```{r load dataset, include = FALSE}
myload(community_metrics, op_analysis, dir = data_dir)
community_metrics %<>% left_join(select(op_analysis, opcod, station, date))
```

```{r}
ts_biomass  <- community_metrics %>%
  select(station, date, biomass) %>%
  mutate(station = as.character(station)) %>%
  group_by(station) %>%
  arrange(date) %>%
  nest()

ts_biomass %<>%
  mutate(
    zts = map(data, function(x) znorm(x$biomass, threshold = 0.01)),
    paa = map(zts, ~paa(., 3)),
    sax = map_chr(paa, ~series_to_string(., 3))
  )
```

```{r}
pattern_dist <- ts_biomass %>%
  group_by(sax) %>%
  summarise(count = n()) %>%
  arrange(desc(count))
ggplot(ts_biomass, aes(x = sax)) +
  geom_histogram(stat = "count")

filter(pattern_dist, count > 10)
```

```{r}
test <- ts_biomass %>%
  filter(sax %in% c("cba", "bbb", "cab", "abc", "acb", "bca", "bac", "cbb")) %>%
  group_by(sax) %>%
  sample_n(10) %>%
  mutate(
    paa2 = map2(data, zts, function (x, zts) {
      cbind(x[,1], zts)
  }),
    id_sta = as.character(seq(1,n()))

  ) %>%
  unnest(paa2)

ggplot(test, aes(x = date, y = zts)) +
  geom_line(aes(color = id_sta)) +
  facet_wrap(~ sax)

```

>>>>

# Assembly/Disassembly of food-web according to changes in species richness and community biomass

### 2025-04-16 paper/appendix.Rmd

Because species biomass are involved in trophic interactions, species losses can
have cascading effects and trigger secondary extinctions [@dunne_network_2002].
In the context of biodiversity crisis, assessing the robustness of ecological
networks to species losses has been then a big question in Ecology. Previous
research showed that food-web structure, such as connectance, increased
robustness [@dunne_network_2002]. The idea is that more connected food-web will
be more stable because they contain a large number of pathways convey biomass
through the food-web. However, connectance as some limitations as a robustness
metric, because each trophic link does not contribute equally to the robustness
of an ecological network [@allesina_who_2004]. Trophic links can be categorised as
functional or redundant, depending on their role in transmitting biomass through
the food-web. Functional links are the set of links that is required to connect
every species to the basal resource of the food-web, all of the other being
redundant. Only functional links affect directly robustness, as their removal
will trigger secondary extinctions. The removal of redundant links in turn do
not directly trigger secondary extinctions, but makes the food-webs more fragile
to further extinctions [@allesina_functional_2009].

The identification of functional and redundant links is based on the use of
denominator trees [@bodini_using_2009; @allesina_who_2004;
@allesina_functional_2009]. Dominator trees allows to identify the set of
trophic links that are essential to connect all the consumers to the basal resource, i.e.
the functional links, and the redundant links, i.e. the redundant trophic pathways
from basal resource to consumers. Pioneer papers explained the algorithm to
compute the set of dominator trees and we refer to them for further
information and examples [@bodini_using_2009; @allesina_who_2004;
@allesina_functional_2009]. Here, we aim to show that while implementing this
analysis could be difficult at the time, this algorithm is now widely
available. In R, it is available in `igraph::dominator_tree()`. Below we show
how the number of functional and redundant links can be  easily computed from R
by replicating three examples presented in previous studies.

@@

Here is an example with the food-web displayed in appendix of
@bodini_using_2009:

@@

@@

We used this function to reproduce the three examples (Fig. \@ref(fig:examples)),
which were displayed in the Fig. 1 and Appendix 1 of @bodini_using_2009, and
in the Fig. 1 of @allesina_functional_2009. The table \@ref(tab:redun) displays
the number and the proportion of functional and redundant links from those
examples.

@@

(ref:example-cap) Reproduced examples of categorising functional links in
food webs using `igraph::dominator_tree()` R function. Left columns (a, c, e)
displays the complete food-web while right column displays food-web with the
functional links only (b, d, f). (a,b): reproduction of Fig. 1 of
@bodini_using_2009, (c, d): reproduction of appendix figures of @bodini_using_2009,
(e, f): reproduction of fig. 1 of @allesina_functional_2009. Note that (e, f) is
a bit different from the original, as noted by the authors, they could have
removed further redundant links (i.e. $2 \to 5$, or $3 \to 5$).

>>>>

Species losses can trigger cascading extinctions by disrupting trophic
interactions [@dunne_network_2002]. To assess ecological network robustness,
researchers often examine food web connectance — the proportion of possible
trophic links that are realized. While higher connectance generally increases
stability by providing multiple pathways for biomass flow, not all links
contribute equally to robustness [@allesina_who_2004].

Trophic links can be categorized into two types based on their role in
maintaining food web integrity. Functional links are those essential for
connecting all species to basal resources — their removal directly triggers
secondary extinctions. In contrast, redundant links provide alternative pathways
but are not strictly necessary for maintaining basic connectivity. While the
loss of redundant links doesn’t cause immediate extinctions, it reduces the
network’s resilience to future perturbations [@allesina_functional_2009]. This
distinction makes quantifying redundancy crucial for predicting ecosystem
vulnerability.

To identify functional and redundant links, we employed dominator trees
[@bodini_using_2009], a graph-theoretic approach that maps the minimal set of
connections required to link all consumers to basal resources. Though once
considered a niche tool, dominator tree algorithms are now readily accessible
through common packages like R’s `igraph::dominator_tree()`. Below, we demonstrate
this method by replicating classic examples from the literature.

For implementation, we used a custom `compute_redundancy` function to quantify
functional and redundant links. As an example, we applied this to the food-web
from Bodini et al. (2009, see Appendix 1 therein), showing how easily these
metrics can be derived from network data.

@@

@@

Figure \@ref(fig:examples) provides visual comparisons between complete food-webs
(left panels) and their functional-link subsets (right panels), reproducing key
results from three seminal studies [@bodini_using_2009;
@allesina_functional_2009]. The accompanying Table \@ref(tab:redun) summarizes
the proportion of redundant links in these systems, highlighting how dominator
trees can simplify robustness assessments.

@@

@@

(ref:example-cap) **Functional link identification using dominator trees**. Left
panels show complete food-webs, while right panels display only functional
links. (a,b) reproduce @bodini_using_2009 Figure 1; (c,d) recreate their
Appendix example; (e,f) replicate @bodini_using_2009 Figure 1 (with minor
differences noted in the original text, they could have
removed further redundant links, i.e. $2 \to 5$, or $3 \to 5$).

### 2025-04-16 paper/method_figure.Rmd

```{r}
library(drake)
library(tidyverse)

@@

  net <- net %>%
    filter(station %in% st) %>%
    dplyr::select(station, year, network) %>%
    unnest(cols = c(network))

  ## Compute metrics:
  net %>%
    dplyr::select(year, from, to) %>%
    arrange(year) %>%
    group_by(year) %>%
    nest(network = c(from, to)) %>%
    mutate(
      igraph_obj = map(network, igraph::graph_from_data_frame,
	directed = TRUE),
      adj_mat = map(igraph_obj, igraph::as_adjacency_matrix,
	sparse = FALSE),
      troph = map(adj_mat, ~NetIndices::TrophInd(.x)),
      obs_troph_level_vector = map(troph, function (x) {
	TL <- x$TL
	names(TL) <- rownames(x) 
	return(TL)
	})
    )
}

@@

```{r}
myload(metaweb_analysis, dir = get_mypath("data"), envir = environment())

my_species_color <- set_color_species(
  node_list = colnames(metaweb_analysis$metaweb),
  resource_list = metaweb_analysis$resource,
  col_resource = NULL
)
```

```{r}
loadd(rigal_classification, com_data, op_data)
op_data %<>%
  mutate(year = as.integer(year))
myload(community_metrics, dir =
  "~/Documents/post-these/mnhn/fishcom_biomass_dynamics/data", envir =
  environment())

community_metrics %<>%
    left_join(dplyr::select(op_data, station, opcod, year))
ti <- rigal_classification %>%
  filter(variable == "bm_std") %>%
  unnest(classif)
first_st <- ti %>%
  filter(shape_class == "decrease_constant") %>%
  arrange(linear_slope)

tes <- map(first_st$station[1:12], ~plot_dyn_web(
    com = community_metrics,
    sem_df = NULL,
    station = .x,
#    .log = TRUE,
    stacked = TRUE
    ))
plot_grid(plotlist = tes)
```
```{r}

last_st <- ti %>%
  filter(shape_class == "increase_constant") %>%
  arrange(linear_slope)

tes <- map(last_st$station[1:12], ~plot_dyn_web(
    com = community_metrics,
    sem_df = NULL,
    station = .x,
#    .log = TRUE,
    stacked = TRUE
    ))
plot_grid(plotlist = tes)
```

```{r}
st <- first_st$station[4]
st[2] <- last_st$station[13]
names(st) <- c("decreasing", "increasing")
#11432 my choice
```

```{r}
myload(network_analysis, dir =
  "~/Documents/post-these/mnhn/fishcom_biomass_dynamics/data/classes", envir =
  environment())
network_analysis %<>%
  left_join(dplyr::select(op_data, opcod, station, year)) %>%
  filter(!is.na(opcod))

p <- get_network_temporal_plot(
  net_list = network_analysis,
  station = st[1],
  selected_years = NULL,
  node_color = NULL, scale_node_size = 0.06,
  par_options = par(mar = c(0,0,0,0)),
  log_abun = TRUE
)
g <- plot_grid(plotlist = map(p, ggdraw), nrow = 3)

save_plot(
  get_mypath("figures/decreasing_biomass_network.pdf"),
  g,
  nrow = 3,
  ncol = 4,
  base_height = 4,
  base_width = 4)

@@

tes <- plot_dyn_web(
    com = mutate(community_metrics, year = ymd(year, truncated = 2L)),
    sem_df = NULL,
    station = st[1],
#    .log = TRUE,
    stacked = TRUE
    )

@@

tes2 <- plot_dyn_web(
    com = mutate(community_metrics, year = ymd(year, truncated = 2L)),
    sem_df = NULL,
    station = st[2],
#    .log = TRUE,
    stacked = TRUE
    )

@@

```{r}
tps_ct_dec <- full_data2 %>%
  mutate(year = ymd(year, truncated = 2L)) %>%
  filter(station == st[1]) %>%
  ggplot(aes(x = year, y = ct_ff)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x,
    size = 1, se = FALSE, color = "black") +
  labs(x = "Year", y = "Connectance")

@@

```{r}
# get a station that represents the results
st_var_shape <- rigal_classification %>%
  unnest(classif) %>%
  filter(variable %in% c(get_com_str_var(), "log_bm_std")) %>%
  select(station, variable, direction)  %>%
  pivot_wider(names_from = variable, values_from = direction)
dec_representative <- st_var_shape %>%
  filter(
    log_bm_std == "decrease",
    ct_ff == "decrease",
    w_trph_lvl_avg == "decrease")
```

```{r}
color_tes <- map(dec_representative$station,
  ~set_color_species(
  node_list = community_metrics %>%
  filter(station == .x) %$%
  .$sp_vector %>%
  reduce(., c) %>%
  names(.) %>%
  unique(),
  resource_list = metaweb_analysis$resource,
  col_resource = NULL
  )
)
names(color_tes) <- as.character(dec_representative$station)

tes <- map(dec_representative$station, ~plot_dyn_web(
    com = community_metrics,
    sem_df = NULL,
    station = .x,
#    .log = TRUE,
    stacked = TRUE,
    color_species = color_tes[[as.character(.x)]]
    ))

names(tes) <- as.character(dec_representative$station)
plot_grid(plotlist = tes)
```
```{r}

st_decreasing <- dec_representative$station[3]

p_dec_log_bm <- tes[[as.character(st_decreasing)]] +
  labs(color = "Species", fill = "Species") +
  theme(legend.position = c(.2, .8), legend.direction = "horizontal")

save_plot(
  get_mypath("figures/p_dec_log_bm.pdf"),
  p_dec_log_bm,
  nrow = 1,
  ncol = 1,
  base_height = 6.6,
  base_asp = 1.2
  )
```
```{r}
v_colors =  viridis(n = 2, begin = 0.5, end = 1)
names(v_colors) <- c("tps", "sp")

tps_dec_inc <- map(c("log_bm_std", "log_rich_std", "ct_ff", "w_trph_lvl_avg", "piel_bm"), # 
  ~plot_temporal_variable_lm(st = st_decreasing, var = .x, col_median =
    v_colors["sp"], col_trend = v_colors["tps"]) +
    theme(axis.title = element_text(size = 12))
)
# Increase point size
tps_dec_inc  <- map(tps_dec_inc, function(x) {
  x$layers[[1]]$aes_params$size <- 4
  return(x)
})

@@

save_plot(
  filename = get_mypath("figures/net_str_dec_bm.pdf"),
  plot_grid(plotlist = tps_dec_inc, nrow = 1, align = "h", axis = "l"),
  ncol = length(tps_dec_inc), nrow = 1, 
  base_height = 2.6,
  base_asp = 1.2 
)
theme_get()$text
```

>>>>

```{r}
library('targets')
library(tidyverse)
library('magrittr')
library(cowplot)

@@

  net <- net %>%
    filter(station %in% st) %>%
    dplyr::select(station, year, network)

  if ("data.frame" %in% class(net$network[[1]])) {

   net <- net %>%
    unnest(cols = c(network)) %>%
    dplyr::select(year, from, to) %>%
    arrange(year) %>%
    group_by(year) %>%
    nest(network = c(from, to)) %>%
    mutate(
      igraph_obj = map(network, igraph::graph_from_data_frame,
	directed = TRUE),
      adj_mat = map(igraph_obj, igraph::as_adjacency_matrix,
	sparse = FALSE),
      troph = map(adj_mat, ~NetIndices::TrophInd(.x)),
      obs_troph_level_vector = map(troph, function (x) {
	TL <- x$TL
	names(TL) <- rownames(x) 
	return(TL)
	})
    )
  } else {

    net <- net %>%
    arrange(year) %>%
    group_by(year) %>%
    rename(adj_mat = network) %>%
    mutate(
      troph = map(adj_mat, ~NetIndices::TrophInd(.x)),
      obs_troph_level_vector = map(troph, function (x) {
	TL <- x$TL
	names(TL) <- rownames(x) 
	return(TL)
	})
    )
  }

  return(net)

@@

```{r}
tar_load(metaweb_analysis)
my_species_color <- set_color_species(
  node_list = colnames(metaweb_analysis$metaweb),
  resource_list = metaweb_analysis$resource,
  col_resource = NULL
)
```

```{r}
tar_load(c(com_data, op_data, com_analysis_data))
op_data %<>%
  mutate(year = as.integer(year))

ti <- com_analysis_data %>%
  group_by(opcod) %>%
  summarise(sp_vector = list(setNames(biomass, species))) %>%
  left_join(dplyr::select(op_data, station, opcod, year, surface))

com_data %<>%
    left_join(dplyr::select(op_data, station, opcod, year))

tar_load(gaussian_pred_station_tps10)
tr10 <- gaussian_pred_station_tps10 %>%
  select(response, station, mean) %>%
  pivot_wider(names_from = "response", values_from = "mean")

first_st <- tr10 %>%
  arrange(log_bm_std)

tes <- map(first_st$station[12:24], ~plot_dyn_web(
    com = ti,
    sem_df = NULL,
    station = .x,
#    .log = TRUE,
    stacked = TRUE
    ))
plot_grid(plotlist = tes)
```
```{r}
last_st <- tr10 %>%
  arrange(desc(log_bm_std))

tes <- map(last_st$station[13:24], ~plot_dyn_web(
    com = ti,
    sem_df = NULL,
    station = .x,
#    .log = TRUE,
    stacked = TRUE
    ))
plot_grid(plotlist = tes)
```

```{r}
st <- first_st$station[20]
st[2] <- last_st$station[16]
names(st) <- c("decreasing", "increasing")
#11432 my choice
```

```{r}
tar_load(c(net_analysis_data, net_data))
load(here::here("data", "classes", "network_analysis.rda"))
network_analysis %<>%
  left_join(dplyr::select(op_data, opcod, station, year)) %>%
  filter(!is.na(opcod))

p <- get_network_temporal_plot(
  net_list = network_analysis,
  station = st[1],
  selected_years = c(1995, 2005),
  node_color = NULL, scale_node_size = 0.02,
  par_options = par(mar = c(0,0,0,0)),
  log_abun = FALSE 
)
g <- plot_grid(plotlist = map(p, ggdraw), ncol = 2)

save_plot(
  get_mypath("figures/decreasing_biomass_network.pdf"),
  g,
  nrow = 1,
  ncol = 2,
  base_height = 4,
  base_width = 4)

@@

tes <- plot_dyn_web(
    com = mutate(ti, year = ymd(year, truncated = 2L)),
    sem_df = NULL,
    station = st[1],
#    .log = TRUE,
    stacked = TRUE
    )

@@

tes2 <- plot_dyn_web(
    com = mutate(com_data, year = ymd(year, truncated = 2L)),
    sem_df = NULL,
    station = st[2],
#    .log = TRUE,
    stacked = TRUE
    )

@@

```{r}
tar_load(full_data2)
tps_ct_dec <- full_data2 %>%
  mutate(year = ymd(year, truncated = 2L)) %>%
  filter(station == st[1]) %>%
  ggplot(aes(x = year, y = ct_ff)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x,
    size = 1, se = FALSE, color = "black") +
  labs(x = "Year", y = "Connectance")

@@

```{r, eval = FALSE}
# get a station that represents the results
st_var_shape <- rigal_classification %>%
  unnest(classif) %>%
  filter(variable %in% c(get_com_str_var(), "log_bm_std")) %>%
  select(station, variable, direction)  %>%
  pivot_wider(names_from = variable, values_from = direction)
dec_representative <- st_var_shape %>%
  filter(
    log_bm_std == "decrease",
    ct_ff == "decrease",
    w_trph_lvl_avg == "decrease")
```

```{r}
color_tes <- map(st[1],
  ~set_color_species(
  node_list = ti %>%
  filter(station == .x) %$%
  .$sp_vector %>%
  reduce(., c) %>%
  names(.) %>%
  unique(),
  resource_list = metaweb_analysis$resource,
  col_resource = NULL
  )
)
names(color_tes) <- as.character(st[1])

tes <- map(st[1], ~plot_dyn_web(
    com = ti,
    sem_df = NULL,
    station = .x,
#    .log = TRUE,
    stacked = TRUE,
    color_species = color_tes[[as.character(.x)]]
    ))

names(tes) <- as.character(st["decreasing"])
plot_grid(plotlist = tes)
```
```{r}
st_decreasing <- st["decreasing"]

p_dec_log_bm <- tes[[as.character(st_decreasing)]] +
  labs(color = "Species", fill = "Species") +
  theme(legend.position = c(.2, .9), legend.direction = "horizontal")

save_plot(
  get_mypath("figures/p_dec_log_bm.pdf"),
  p_dec_log_bm,
  nrow = 1,
  ncol = 1,
  base_height = 4,
  base_width = 7
  )
```

```{r}
tar_load(c(net_analysis_data, net_data))
load(here::here("data", "classes", "network_analysis.rda"))
network_analysis %<>%
  left_join(dplyr::select(op_data, opcod, station, year)) %>%
  filter(!is.na(opcod))

p <- get_network_temporal_plot(
  net_list = network_analysis,
  station = st[1],
  selected_years = c(1995, 2005),
  node_color = color_tes[[st[1]]], scale_node_size = 0.05,
  par_options = par(mar = c(0, 0, 0, 0)),
  log_abun = TRUE 
)
g <- plot_grid(plotlist = map(p, ggdraw), ncol = 2)

save_plot(
  get_mypath("figures/decreasing_biomass_network.pdf"),
  g,
  nrow = 1,
  ncol = 2,
  base_height = 4,
  base_width = 4)

```{r}
v_colors =  viridisLite::viridis(n = 2, begin = 0.5, end = 1)
names(v_colors) <- c("tps", "sp")

tps_dec_inc <- map(c("log_bm_std", "log_rich_std", "ct_ff", "w_trph_lvl_avg"), # 
  ~plot_temporal_variable_lm(st = st_decreasing, var = .x, col_median =
    v_colors["sp"], col_trend = v_colors["tps"]) +
    theme(axis.title = element_text(size = 12))
)
# Increase point size
tps_dec_inc  <- map(tps_dec_inc, function(x) {
  x$layers[[1]]$aes_params$size <- 4
  return(x)
})

@@

save_plot(
  filename = get_mypath("figures/net_str_dec_bm.pdf"),
  plot_grid(plotlist = tps_dec_inc, nrow = 1, align = "h", axis = "l"),
  ncol = length(tps_dec_inc), nrow = 1, 
  base_height = 2.6,
  base_asp = 1.2 
)
```
